\LARGE \bf IMU-based pose-estimation for spherical robots with limited resources

### 2021-05-29 mfi21/ieeeconf/root.tex
small changes

\overrideIEEEmargins                                      % Needed to meet printer requirements.
\usepackage{graphicx}
\usepackage{amsmath}
%\usepackage{enumitem}
\newcommand{\M}[1]{\mathbf{#1}} % Bold for matrix
\newcommand{\V}[1]{\mathbf{#1}} % Bold for vector
%In case you encounter the following error:
%Error 1010 The PDF file may be corrupt (unable to open PDF file) OR
%Error 1000 An error occurred while parsing a contents stream. Unable to analyze the PDF file.
%This is a known problem with pdfLaTeX conversion filter. The file cannot be opened with acrobat reader
%Please use one of the alternatives below to circumvent this error by uncommenting one or the other
%\pdfobjcompresslevel=0
%\pdfminorversion=4

@@

Spherical robots are a relatively narrow field of robotics. Still, they could be useful in situations where the measurement equipment needs to be protected against harsh environments.
As an example, the 2021 CDF study LunarCaves by ESA about the feasibility of a spherical robot, called DAEDALUS, for exploration of lunar-lava tubes \cite{c2} shows the need for orientation and position estimation of spherical robots with the limitations on resources common for space-applications.
Although a purely Inertial Measurement Unit (IMU)-based estimation is not recommended for various reasons, the lack of an absolute reference among others, IMU-based pose estimation are used as part of an overall multi-sensor-fusion-based estimation.
IMUs are sensors with higher refresh rates than Simultaneous Localization and Mapping (SLAM) based algorithms, such as optical or Light Detection and Ranging (LIDAR) SLAM. This makes them suitable for interpolating the more precise and absolute position reference points of these slower algorithms.
The limited computational power of space-qualified CPUs further imposes the need for good orientation estimation without complex algorithms like Kalman filters.
Lastly, the geometry of a spherical robot opens the possibility for the not often used position estimation by IMUs.

@@

\paragraph*{Real Time Operations}The operation of the robot does not allow post-calculation of the pose. This implies the usage of the onboard computational power.
\paragraph*{Limitation of Computational Power}
The need for efficient code without massive bottleneck operations, as the inversion of a matrix with the extended Kalman Filter. The code must minimize the usage of non-basic operations.
\paragraph*{Low Costs of Hardware}
We assume biased and noisy data in a non-negligible amount from the sensors.
\paragraph*{Sphere Shape}
The code relies heavily on the assumption of a sphere as a shape with a constant robot radius. We also assume that rotation of the sphere on the ground leads to translation.
\paragraph*{Slip and Sliding}
We assume the sphere has both slip, i.e., the sphere rotates, but there is no translation, and sliding, i.e., the sphere has translation without rotation in a limited, non-permanent way. The limitation is the assumption that sliding and slipping reduce the amount of translation for a given rotation and vice versa, however, not for complete absence.
\paragraph*{Stability of Pose}
For simultaneous localization and mapping (SLAM) purposes using LIDARs, a stable but maybe minimal false pose is preferred over a more exact but noisy and jumping position.
\paragraph*{Further Processing of Pose}
The approach should avoid jumping values or abrupt changes of data if not clearly indicated by the sensors. Suppose there is an internal algorithmic change of behavior due to the change of state from standing to rolling. This change shall not be represented by an unnatural acceleration in the data.
\paragraph*{Uncertainty of Iterative Position Integration}
IMUs deliver no absolute reference for a position as they do for orientation in the form of the noisy but measurable gravity vector. So without external or secondary sensors relying on absolute waypoints, the position estimation will always underlie the integration of errors.
\paragraph*{Space Suitability}
As the code is considered to be applied to space missions, the magnetometer data is not considered in the algorithm, despite its positive impact on orientation estimation on earth.
\paragraph*{Short Term Mobility}
As the code still relies on integration parts, the sphere is not considered to have a straight motion for a long period but rather makes multiple short paths. This will also be necessary, as in further extension, the fusion of a pose estimation by the laser scanner will be merged. 
\paragraph*{Non-reliablity of locomotion}
The sphere from the DAEDALUS example has two completely different locomotion methods; therefore, we will not take locomotion itself into account of pose estimation. Also, the same locomotion approach leads to heavily different behavior with the same input to the controller~\cite{c2}.
\paragraph*{Classification of Data}
The generated data is a compromise between simple filtration of data and heavy calculation post-record. It is easy to calculate and adapt, but due to the number of pre-assumptions, as it makes assumptions about the dynamics of the system, it still delivers reliable data which takes physical behavior into account. It is real-time capable and does not rely on delay steps.
\paragraph*{Field of Use}
The use is restricted to spherical or cylindrical robots. It is optimized for gathering laser data where noise position while standing means a lot of calculation in the map creation, and therefore, even if slightly false, a stable position is preferred.

\section{State of the Art}
\label{sec:SOA}
For pose estimation on embedded hardware, there are three main algorithms.
\subsection{Kalman Filter}
The Kalman Filter or the more suitable Extended Kalman Filter (EKF) uses state prediction and merges with sensor data for orientation estimation.
For the defined limitations, the Kalman Filter is not suitable.
First, the EKF requires a considerable amount of computational power, mainly because of its matrix inversion. For this, the needed calculations grow exponentially for input data.
Further, the Kalman Filter relies on input to a system to calculate an estimation of the pose for the next iterative step.
But, as we described in the preliminary section, we can not take the input to locomotion control as a reference to pose estimation as the behavior is not linear or even predictable without exact knowledge of the surface.
Therefore it is not possible to implement a physical behavior of the system to the filter, as easily as it is for other robotic applications.
\subsection{Madgwick Filter}
The Madgwick filter is a widely used filter for orientation estimation with IMUs~\cite{c1}.
It is efficient and easy to calculate but does not consider the physics of the motion.
It does nearly provide the sought-after classification for the orientation.
The problem is the optimization for movement. The Madgwick filter has a jittering behavior at the standstill of the IMU.
This does not meet the desired behavior as described in the prelimitations section (\ref{AssumptionsAndPrelimitations}).
The behavior is shown in comparison to the complementary filter in Figure~\ref{madgVsComp1}.
This behavior is most prominent during standstill as the Madgwick filter was designed for movement.
\subsection{Complementary Filter}
A complementary filter is a fundamental approach of combining precise but not exact with nonprecise but exact data.
Therefore it is widely used for combining noisy accelerometer data with gravity as a reference with the precise but drifting measurement of gyroscopes.
The basic concept is to have the main part of a value dominated by a gyroscope but, over time, a slow drift towards the measurement of the accelerometer.
E.g., for pitch, this would result in 
\begin{equation}
\theta_n = (1-\alpha) \cdot (\theta_{n-1}+ \Delta\theta_{gyro})+ \alpha \cdot \theta_{acc}
\end{equation}
where $\alpha$ is the complementary gain, it determines how much weight the accelerometer data gets and how fast the value drifts towards the measurement of the accelerometer.
A commonly used value for $\alpha$ is 0.02, meaning every iteration, 2 percent of the new values come from the accelerometer.
An $\alpha$ of 1 means the resulting data is completely the one of the accelerometers, and the gyroscope has no influence. \\
The complementary filter is best for orientation calculation without position change as the accelerometer value relies on calculating the gravity value.
With an acceleration in one direction, this leads to a miscalculation of the gravitational vector and, therefore, a false orientation estimation.
Simple orientation estimations of quadrocopters often just change the $\alpha$ gain to 0 while moving to avoid the miscalculation of the gravity vector \cite{complementary}.

@@

\subsection{Position Estimation}
Position estimation purely by IMU is not recommended as the double integration of the accelerometer measurements leads to exponential error summation.
Therefore there are very few approaches targeting this way of position calculating \cite{c3}.
The very impressive work in \cite{Valencia2017ASM} also uses IMUs for 6D pose estimation.
They follow a pre-specified trajectory with a manipulator and manage to estimate its pose within 10\% of the ground truth.
Given their application is rather broad, they do not exploit known dynamics of the system, which could be helpful for spherical robots.

@@

\subsection{Orientation}
The orientation is a symbiosis of the Madgwick filter and complementary filter. For no- or slow-moving, the complementary filter is chosen; for fast-moving, the Madgwick filter. The Madgwick filter itself is untouched.
The standard complimentary filter is adapted to work in quaternion representation.
 As the integration only works in the Euler angles domain, the jumps of values at the border of the domain like 0 and 360 degrees do not match the mostly continuous change of quaternions.
As the representation is still valid (0 degree and 360-degree roll being the same position), this is not a problem when using at one point any of the filters exclusively.
 But combining both leads to contradicting updates from each filter with respect to the other.
 Therefore the normal way of the complementary filter of the new orientation is 
%\begin{equation}
%\begin{bmatrix}
%\Phi_{new} \\
%\theta_{new}\\
%\Psi_{new}
%\end{bmatrix}
%= (1-\alpha)\begin{bmatrix}
%\Phi_{old} \\
%\theta_{old}\\
%\Psi_{old}
%\end{bmatrix}
% + \alpha \begin{bmatrix}
%\Phi_{acc} \\
%\theta_{acc}\\
%\Psi_{acc}
%\end{bmatrix}
%\end{equation}
not used, as it implies a transition of each value independently.
Hence, the RPY-orientation is transformed to a quaternion, and then a quaternion slerp is performed \cite{c4}.
 The quaternion slerp works over the representation of two quaternions on a sphere. The transition from one quaternion to the other is now not done directly but always over the surface of the sphere.
 This leads to the same handling of quaternions as the Madgwick-filter without endangering robustness due to using two different quaternions with very different values representing nearly the same position.

@@

\subsection{Orientation}
A first experiment is performed to show the overall performance of the Autogain computation for the orientation. The IMU was manually rotated in arbitrary directions, and the orientation plotted using the Madgwick filter and the Autogain filter.
As expected, the overall behavior of the presented algorithm and the Madgwick filter is very similar in the dynamic case, as shown in Figure~\ref{quaternion1}.
\begin{figure}
\includegraphics[width=\linewidth]{./graphics/AutoVsMad.eps}
\caption{Quaternion-values of the presented fusion of complementary and Madgwick filter and only the Madgwick filter}
\label{quaternion1}
\end{figure}
Therefore the more expressive part is the change between motion and standstill, as in motion, both filters use the same algorithm but differ during standstill and slow motion.
As described in Section \ref{sec:SOA}, the Madgwick filter has a rather strong jitter during slow movements and standstill. 
To evaluate specifically the change between slow motion and standstill, a further experiment is performed.
Figure \ref{distance} show the distance to a precise external reference over a short excerpt of the experiment with three phases of slow motion and rotation at second 7, 9 and 11. 
For external reference, we used a precise optical tracking system. 
The definition and proposed calculation of the distance of 2 quaternions of \cite{kuffner} was used.
It describes the direct angle between two quaternions.
The optical system matches the overall orientation of both IMU-based algorithms quite well and lacks the perturbations of the pure Madgwick filter.
The transformation of the orientation by the optical tracking system and the IMU-based algorithms yield two unknown transformations, the relative orientation of the coordinate system of IMU to optical coordinate system and how the axes are interpreted on the tracked structure itself.
Therefore, the algorithm proposed in \cite{dornaika} was used to solve these two unknown transformations on basis of the resulting data, to match the orientations of optical and IMU-based data.
As this matching of the optical system happens on basis of one of the two IMU orientation data-sets, the comparison of absolute differences may be misleading, as the transformations were matched on basis of the autogain filter. 
To stress the jitter behavior of the Madgwick without relying on absolute orientation, which will be part of further research, Figure \ref{diffdistance} shows the derivative of the distance.
Here the change of direction due to small oscillations and the smoother fusion algorithm in comparison to the tracking is even more apparent.

\begin{figure}
\includegraphics[width=\linewidth]{./graphics/distance.eps}
\caption{Distance of the proposed and the Madgwick algorithm to the optical tracking system, using the distance definiton of \cite{kuffner}}
\label{distance}
\end{figure}

\begin{figure}
\includegraphics[width=\linewidth]{./graphics/diffdistance.eps}
\caption{Derivative of the distance of the proposed and the Madgwick algorithm to the optical tracking system, using the distance definiton of \cite{kuffner}}
\label{diffdistance}
\end{figure}

@@

\subsection{Conclusion}
This paper introduced an IMU-based pose estimation filter based on the fusion of multiple already existing filters and approaches.
It is optimized for specific needs and circumstances of spherical robots with laser scanning tasks in a space-suitable way.
Needless to say, a lot of work remains to be done.
The fusion of complementary and Madgwick orientation filter solves the shortcomings of the Madgwick when at a standstill, particularly the jitter.
A first qualitative evaluation shows the functionality of the filter.
In further experiments, a quantitative Orientation comparison with a synchronized optical system needs to be performed.
The position estimation, as well as the combination of the integration by rotation of the gyroscope and the linear acceleration of the accelerometer, shows the capability to overcome the problem of the missing absolute reference of IMUs partly, which mainly improves the measurements in the ground plane with slip behavior present.
Still, there is a need for parameter tuning. 
Sliding was not detected all the time reliably.
Here a mechanism for a delayed rotation needs to be investigated, that acceleration shortly before a rotation is integrated and taken into account once rotation starts.
The approach for, at least roughly, indicating the change in the $z$-direction, which is now done only by the accelerometer values, shows no promising results nor usable results in first qualitative experiments with rolling over obstacles and therefore needs to be re-investigated.

@@

\begin{thebibliography}{99}
\bibitem{c2} Rossi, A.P. et al., "DAEDALUS - Descent And Exploration in Deep Autonomy of Lava Underground Structures", Wuerzburger Forschungsberichte in Robotik und Telematik (21), 978-3-945459-33-1, (2021).
\bibitem{c1} Madgwick, Sebastian O. H., "An efficient orientation filter for inertial and inertial / magnetic sensor arrays", (2010).
\bibitem{complementary}Mahony, R. ; Hamel, T.; Pflimlin, J.-M.,"Complementary filter design on the special orthogonal group SO(3)", Proceedings of the 44th IEEE Conference on Decision and Control, (2005).
\bibitem{c3} Kok, Manon; Hol, Jeroen D.;  Schoen,Thomas B.,  "Using Inertial Sensors for Position and Orientation Estimation", Foundations and Trends in Signal Processing: Vol. 11.(2017).
\bibitem{Valencia2017ASM} Botero Valencia, J.-S. ; Garc{\'i}a, M. R. ; Ceballos, J. V., "A simple method to estimate the trajectory of a low cost mobile robotic platform using an IMU", International Journal on Interactive Design and Manufacturing (ijidem), (2017).
\bibitem{c4}Dam, Erik B.; Koch, Martin;  Lillholm, Martin, "Quaternions, Interpolation and Animation", (1998).
\bibitem{FastInvSqrt}"Origin of Quake3's Fast InvSqrt()", beyond3d.com/content/articles/8/, Accessed: 2021-05-01.
\bibitem {kuffner}Kuffner, James, "Effective sampling and distance metrics for 3D rigid body path planning", Proceedings - IEEE International Conference on Robotics and Automation, pp. 3993 - 3998 Vol.4, (2004).
\bibitem{dornaika}Dornaika, F.; Horaud, R., Simultaneous robot-world and hand-eye calibration, IEEE Transactions on Robotics and Automation 14(4), pp. 617-622, (1998).

\end{thebibliography}

>>>>

\overrideIEEEmargins                                      % Needed to meet printer requirements.
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{hyperref}
%\usepackage{enumitem}
\newcommand{\M}[1]{\mathbf{#1}} % Bold for matrix
\newcommand{\V}[1]{\mathbf{#1}} % Bold for vector
%In case you encounter the following error:
%Error 1010 The PDF file may be corrupt (unable to open PDF file) OR
%Error 1000 An error occurred while parsing a contents stream. Unable to analyze the PDF file.
%This is a known problem with pdfLaTeX conversion filter. The file cannot be opened with acrobat reader
%Please use one of the alternatives below to circumvent this error by uncommenting one or the other
%\pdfobjcompresslevel=0
%\pdfminorversion=4

@@

Spherical robots are a relatively narrow field of robotics. Still, they could be useful in situations where the measurement equipment needs to be protected against harsh environments.
As an example, the 2021 CDF study LunarCaves by ESA about the feasibility of a spherical robot, called DAEDALUS, for exploration of lunar-lava tubes \cite{rossi2021daedalus} shows the need for orientation and position estimation of spherical robots with the limitations on resources common for space-applications.
Although a purely Inertial Measurement Unit (IMU)-based estimation is not recommended for various reasons, the lack of an absolute reference among others, IMU-based pose estimation are used as part of an overall multi-sensor-fusion-based estimation.
IMUs are sensors with higher refresh rates than Simultaneous Localization and Mapping (SLAM) based algorithms, such as optical or Light Detection and Ranging (LIDAR) SLAM. This makes them suitable for interpolating the more precise and absolute position reference points of these slower algorithms.
The limited computational power of space-qualified CPUs further imposes the need for good orientation estimation without complex algorithms like Kalman filters.
Lastly, the geometry of a spherical robot opens the possibility for the not often used position estimation by IMUs.

@@

\paragraph*{Real Time Operations}The operation of the robot does not allow post-calculation of the pose. This implies the usage of the onboard computational power.
\paragraph*{Limitation of Computational Power}
The need for efficient code without massive bottleneck operations, as the inversion of a matrix with the extended Kalman Filter. The code must minimize the usage of non-basic operations.
\paragraph*{Low Costs of Hardware}
We assume biased and noisy data in a non-negligible amount from the sensors.
\paragraph*{Sphere Shape}
The code relies heavily on the assumption of a sphere as a shape with a constant robot radius. We also assume that rotation of the sphere on the ground leads to translation.
\paragraph*{Slip and Sliding}
We assume the sphere has both slip, i.e., the sphere rotates, but there is no translation, and sliding, i.e., the sphere has translation without rotation in a limited, non-permanent way. The limitation is the assumption that sliding and slipping reduce the amount of translation for a given rotation and vice versa, however, not for complete absence.
\paragraph*{Stability of Pose}
For simultaneous localization and mapping (SLAM) purposes using LIDARs, a stable but maybe minimal false pose is preferred over a more exact but noisy and jumping position.
\paragraph*{Further Processing of Pose}
The approach should avoid jumping values or abrupt changes of data if not clearly indicated by the sensors. Suppose there is an internal algorithmic change of behavior due to the change of state from standing to rolling. This change shall not be represented by an unnatural acceleration in the data.
\paragraph*{Uncertainty of Iterative Position Integration}
IMUs deliver no absolute reference for a position as they do for orientation in the form of the noisy but measurable gravity vector. So without external or secondary sensors relying on absolute waypoints, the position estimation will always underlie the integration of errors.
\paragraph*{Space Suitability}
As the code is considered to be applied to space missions, the magnetometer data is not considered in the algorithm, despite its positive impact on orientation estimation on earth.
\paragraph*{Short Term Mobility}
As the code still relies on integration parts, the sphere is not considered to have a straight motion for a long period but rather makes multiple short paths. This will also be necessary, as in further extension, the fusion of a pose estimation by the laser scanner will be merged. 
\paragraph*{Non-reliablity of locomotion}
The sphere from the DAEDALUS example has two completely different locomotion methods; therefore, we will not take locomotion itself into account of pose estimation. Also, the same locomotion approach leads to heavily different behavior with the same input to the controller \cite{rossi2021daedalus}.
\paragraph*{Classification of Data}
The generated data is a compromise between simple filtration of data and heavy calculation post-record. It is easy to calculate and adapt, but due to the number of pre-assumptions, as it makes assumptions about the dynamics of the system, it still delivers reliable data which takes physical behavior into account. It is real-time capable and does not rely on delay steps.
\paragraph*{Field of Use}
The use is restricted to spherical or cylindrical robots. It is optimized for gathering laser data where noise position while standing means a lot of calculation in the map creation, and therefore, even if slightly false, a stable position is preferred.

\section{State of the Art}
\label{sec:SOA}
For pose estimation on embedded hardware, there are three main algorithms.
\subsection{Kalman Filter}
The Kalman Filter or the more suitable Extended Kalman Filter (EKF) uses state prediction and merges with sensor data for orientation estimation \cite{Kalman1960}.
For the defined limitations, the Kalman Filter is not suitable.
First, the EKF requires a considerable amount of computational power, mainly because of its matrix inversion. For this, the needed calculations grow exponentially for input data.
Further, the Kalman Filter relies on input to a system to calculate an estimation of the pose for the next iterative step.
But, as we described in the preliminary section, we can not take the input to locomotion control as a reference to pose estimation as the behavior is not linear or even predictable without exact knowledge of the surface.
Therefore it is not possible to implement a physical behavior of the system to the filter, as easily as it is for other robotic applications.
\subsection{Madgwick Filter}
The Madgwick filter is a widely used filter for orientation estimation with IMUs \cite{madgwick2010}.
It is efficient and easy to calculate but does not consider the physics of the motion.
It does nearly provide the sought-after classification for the orientation.
The problem is the optimization for movement. The Madgwick filter has a jittering behavior at the standstill of the IMU.
This does not meet the desired behavior as described in the prelimitations section (\ref{AssumptionsAndPrelimitations}).
The behavior is shown in comparison to the complementary filter in Figure~\ref{madgVsComp1}.
This behavior is most prominent during standstill as the Madgwick filter was designed for movement.
\subsection{Complementary Filter}
A complementary filter is a fundamental approach of combining precise but not exact with nonprecise but exact data.
Therefore it is widely used for combining noisy accelerometer data with gravity as a reference with the precise but drifting measurement of gyroscopes.
The basic concept is to have the main part of a value dominated by a gyroscope but, over time, a slow drift towards the measurement of the accelerometer.
E.g., for pitch, this would result in 
\begin{equation}
\theta_n = (1-\alpha) \cdot (\theta_{n-1}+ \Delta\theta_{gyro})+ \alpha \cdot \theta_{acc}
\end{equation}
where $\alpha$ is the complementary gain, it determines how much weight the accelerometer data gets and how fast the value drifts towards the measurement of the accelerometer.
A commonly used value for $\alpha$ is 0.02, meaning every iteration, 2 percent of the new values come from the accelerometer.
An $\alpha$ of 1 means the resulting data is completely the one of the accelerometers, and the gyroscope has no influence. \\
The complementary filter is best for orientation calculation without position change as the accelerometer value relies on calculating the gravity value.
With an acceleration in one direction, this leads to a miscalculation of the gravitational vector and, therefore, a false orientation estimation.
Simple orientation estimations of quadrocopters often just change the $\alpha$ gain to 0 while moving to avoid the miscalculation of the gravity vector \cite{mahony2005complementary}.

@@

\subsection{Position Estimation}
Position estimation purely by IMU is not recommended as the double integration of the accelerometer measurements leads to exponential error summation.
Therefore there are very few approaches targeting this way of position calculating \cite{kok2017IMU}.
The very impressive work in \cite{valencia2017simpleIMU} also uses IMUs for 6D pose estimation.
They follow a pre-specified trajectory with a manipulator and manage to estimate its pose within 10\% of the ground truth.
Given their application is rather broad, they do not exploit known dynamics of the system, which could be helpful for spherical robots.

@@

\subsection{Orientation}
The orientation is a symbiosis of the Madgwick filter and complementary filter. For no- or slow-moving, the complementary filter is chosen; for fast-moving, the Madgwick filter. The Madgwick filter itself is untouched.
The standard complimentary filter is adapted to work in quaternion representation.
 As the integration only works in the Euler angles domain, the jumps of values at the border of the domain like 0 and 360 degrees do not match the mostly continuous change of quaternions.
As the representation is still valid (0 degree and 360-degree roll being the same position), this is not a problem when using at one point any of the filters exclusively.
 But combining both leads to contradicting updates from each filter with respect to the other.
 Therefore the normal way of the complementary filter of the new orientation is 
%\begin{equation}
%\begin{bmatrix}
%\Phi_{new} \\
%\theta_{new}\\
%\Psi_{new}
%\end{bmatrix}
%= (1-\alpha)\begin{bmatrix}
%\Phi_{old} \\
%\theta_{old}\\
%\Psi_{old}
%\end{bmatrix}
% + \alpha \begin{bmatrix}
%\Phi_{acc} \\
%\theta_{acc}\\
%\Psi_{acc}
%\end{bmatrix}
%\end{equation}
not used, as it implies a transition of each value independently.
Hence, the RPY-orientation is transformed to a quaternion, and then a quaternion slerp is performed \cite{dam1998quaternions}.
 The quaternion slerp works over the representation of two quaternions on a sphere. The transition from one quaternion to the other is now not done directly but always over the surface of the sphere.
 This leads to the same handling of quaternions as the Madgwick-filter without endangering robustness due to using two different quaternions with very different values representing nearly the same position.

@@

\subsection{Orientation}
A first experiment is performed to show the overall performance of the Autogain computation for the orientation. The IMU was manually rotated in arbitrary directions, and the orientation plotted using the Madgwick filter and the Autogain filter.
As expected, the overall behavior of the presented algorithm and the Madgwick filter is very similar in the dynamic case, as shown in Figure \ref{quaternion1}.
\begin{figure}
\includegraphics[width=\linewidth]{./graphics/AutoVsMad.eps}
\caption{Quaternion-values of the presented fusion of complementary and Madgwick filter and only the Madgwick filter}
\label{quaternion1}
\end{figure}
Therefore the more expressive part is the change between motion and standstill, as in motion, both filters use the same algorithm but differ during standstill and slow motion.
As described in Section \ref{sec:SOA}, the Madgwick filter has a rather strong jitter during slow movements and standstill. 
To evaluate specifically the change between slow motion and standstill, a further experiment is performed.
Figure \ref{distance} show the distance to a precise external reference over a short excerpt of the experiment with three phases of slow motion and rotation at second 7, 9 and 11. 
For external reference, we used a precise optical tracking system. 
The definition and proposed calculation of the distance of 2 quaternions of \cite{kuffner2004effectiveDistance} was used.
It describes the direct angle between two quaternions.
The optical system matches the overall orientation of both IMU-based algorithms quite well and lacks the perturbations of the pure Madgwick filter.
The transformation of the orientation by the optical tracking system and the IMU-based algorithms yield two unknown transformations, the relative orientation of the coordinate system of IMU to optical coordinate system and how the axes are interpreted on the tracked structure itself.
Therefore, the algorithm proposed in \cite{dornaika1998calibration} was used to solve these two unknown transformations on basis of the resulting data, to match the orientations of optical and IMU-based data.
As this matching of the optical system happens on basis of one of the two IMU orientation data-sets, the comparison of absolute differences may be misleading, as the transformations were matched on basis of the autogain filter. 
To stress the jitter behavior of the Madgwick without relying on absolute orientation, which will be part of further research, Figure \ref{diffdistance} shows the derivative of the distance.
Here the change of direction due to small oscillations and the smoother fusion algorithm in comparison to the tracking is even more apparent.

\begin{figure}
\includegraphics[width=\linewidth]{./graphics/distance.eps}
\caption{Distance of the proposed and the Madgwick algorithm to the optical tracking system, using the distance definiton of \cite{kuffner2004effectiveDistance}}
\label{distance}
\end{figure}

\begin{figure}
\includegraphics[width=\linewidth]{./graphics/diffdistance.eps}
\caption{Derivative of the distance of the proposed and the Madgwick algorithm to the optical tracking system, using the distance definiton of \cite{kuffner2004effectiveDistance}}
\label{diffdistance}
\end{figure}

@@

\subsection{Conclusion}
This paper introduced an IMU-based pose estimation filter based on the fusion of multiple already existing filters and approaches.
It is optimized for specific needs and circumstances of spherical robots with laser scanning tasks in a space-suitable way.
Needless to say, a lot of work remains to be done.
The fusion of complementary and Madgwick orientation filter solves the shortcomings of the Madgwick when at a standstill, particularly the jitter.
A first qualitative evaluation shows the functionality of the filter.
In further experiments, a quantitative Orientation comparison with a synchronized optical system needs to be performed.
Also as further work, all puzzle pieces, this filter, a locomotion system, a scanning system,  need to be combined to an overall working prototypes.
The position estimation, as well as the combination of the integration by rotation of the gyroscope and the linear acceleration of the accelerometer, shows the capability to overcome the problem of the missing absolute reference of IMUs partly, which mainly improves the measurements in the ground plane with slip behavior present.
Still, there is a need for parameter tuning. 
Sliding was not detected all the time reliably.
Here a mechanism for a delayed rotation needs to be investigated, that acceleration shortly before a rotation is integrated and taken into account once rotation starts.
The approach for, at least roughly, indicating the change in the $z$-direction, which is now done only by the accelerometer values, shows no promising results nor usable results in first qualitative experiments with rolling over obstacles and therefore needs to be re-investigated.

@@

\bibliographystyle{unsrt}
\bibliography{bibliographyJasper}
